Comparing microbial communities with OptiFit: a fast method for fitting sequences to existing OTUs

### 2021-03-11 docs/slides/2021-03-11_bistro.Rmd
Minor tweaks

@@

1. Calculate all pairwise sequence distance scores.

@@

.center[
<img src="../../figures/opticlust-supp-ex.png" width="700"/>
]
.footnote[
<img src="https://raw.githubusercontent.com/mothur/logo/master/mothur_RGB.png" width="200"/>
]
???
if equally good, pick one at random

@@

- Modes
  - Open - sequences that can't be fit to the reference are then clustered _de novo_ to create additional OTUs.
  - Closed - sequences that can't be fit to the reference are thrown out.

- Reference choice
  - External, independent database (Silva, greengenes, RDP).
  - Split the dataset into a reference & query set.

@@

## Sequence similarity / distance

1. Calculate all pairwise sequence identity scores to create a distance matrix.
2. Apply cutoff of 97% sequence identity to create an adjacency matrix.
.center[
```
seqA AGGGTACG
seqB ACG-TACG
```
$$ \%\ sequence\ identity =  100 * \frac{number\ of\ matches}{sequence\  length} $$
]
.pull-left[
1) Distance matrix:

| | Seq1 | Seq2 | Seq3 |
|-|------|------|------|
| Seq1 | - | 98 | 87 |
| Seq2 | 98 | - | 76 |
| Seq3 | 87 | 76 | - |
]
.pull-right[
2) Adjacency matrix:

| | Seq1 | Seq2 | Seq3 |
|-|------|------|------|
| Seq1 | - | 1 | 0 |
| Seq2 | 1 | - | 0 |
| Seq3 | 0 | 0 | - |
]

>>>>

???
Sarah Westcott, software engineer.
implemented algorithms I'm discussing today.

@@

1. Calculate all pairwise distance scores.

  .center[
```
seqA AGGGTACG
seqB ACG-TACG
```
$$distance = 1 - sequence\ similarity = 1 - \frac{number\ of\ matches}{sequence\  length}$$
]

@@

.center[
<img src="../../figures/opticlust-supp-ex.png" width="700"/>
]

.footnote[
<img src="https://raw.githubusercontent.com/mothur/logo/master/mothur_RGB.png" height="150"/>
]

???
if equally good, pick one at random.

opticlust is great, but still has limitations of _de novo_ clustering: OTU assignments aren't stable when new sequences are added.

@@

- Modes
  - Closed - sequences that can't be fit to the reference are thrown out.
  - Open - sequences that can't be fit to the reference are then clustered _de novo_ to create additional OTUs.

- Reference choice
  - External, independent database (SILVA, greengenes, RDP).
  - Split the dataset into a reference & query set.

@@
